Early Observations of the Type Ia Supernova \abc:\\ A Case of Interaction with Nearby, Unbound Material and/or Strong Ejecta Mixing

### 2018-05-22 doc/ms.tex

%% Reintroduced the \received and \accepted commands from AASTeX v5.2
\received{\today}
\revised{}
\accepted{}
%% Command to document which AAS Journal the manuscript was submitted to.
%% Adds "Submitted to " the arguement.
\submitjournal{ApJ}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% The following section outlines numerous optional output that
%% can be displayed in the front matter or as running meta-data.
%%
%% If you wish, you may supply running head information, although
%% this information may be modified by the editorial offices.
\shorttitle{\abc}
\shortauthors{Miller et al.}
%%
%% You can add a light gray and diagonal water-mark to the first page 
%% with this command:
\watermark{DRAFT}
%% where "text", e.g. DRAFT, is the text to appear.  If the text is 
%% long you can control the water-mark size with:
%  \setwatermarkfontsize{dimension}
%% where dimension is any recognized LaTeX dimension, e.g. pt, in, etc.
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Early observations of Type Ia supernovae (SNe Ia) provide a unique probe of
their progenitor systems and explosion physics. Here we report the
intermediate Palomar Transient Factory (iPTF) discovery of an
extraordinarily young SN Ia, \abc. By fitting a power law to our early light
curve, we infer that first light for the SN, that is when the SN could have
first been detected by our survey, occurred only
$0.15\pm_{0.07}^{0.15}$\,days before our first detection. In the
$\sim$24\,hr after discovery, \abc\ rose by $\sim$2\,mag, featuring a
near-linear rise in flux for $\gtrsim$3\,days. Early spectra show strong
\ion{C}{2} absorption, which disappears after $\sim$7\,days. Unlike the
extensivelyobserved SN Ia SN\,2011fe, the $(B-V)_0$ colors of \abc\ are blue
and nearly constant in the days after explosion. We show that our early
observations of \abc\ cannot be explained by either SN shock breakout and
the associated, subsequent cooling or the SN ejecta colliding with a stellar
companion. Instead, we argue that the early characteristics of \abc,
including (i) the rapid, near-linear rise, (ii) the nonevolving blue colors,
and (iii) the strong \ion{C{2}} absorption, are the result of either ejecta
interaction with nearby, unbound material or vigorous mixing of radioactive
$^{56}$Ni in the SN ejecta, or a combination of the two. In the next few
years, dozens of very young \textit{normal} SNe Ia will be discovered, and
observations similar to those presented here will constrain the white dwarf
explosion mechanism.

@@

The vast majority of SNe Ia are observed to be powered purely by the
radioactive decay of $^{56}$Ni. While the detection of SN shock cooling or
ejecta interaction is rare, the level of $^{56}$Ni mixing in the SN ejecta
can fundamentally alter the appearance of the SN shortly after explosion
(e.g.,
\citealt{2014MNRAS.441..532D,2016ApJ...826...96P,2017MNRAS.472.2787N)}.

SNe Ia experience a dark phase after the SN shock breakout but before
radioactive energy diffuses into the photosphere
\citep{2014ApJ...784...85P}. The duration of this dark phase is set by how
the newly synthesized $^{56}$Ni is mixed and deposited into different layers
of the ejecta. Strong mixing leads to a short, or nonexistent, dark phase,
because the radioactive $\gamma$-rays rapidly diffuse to the photosphere.
This also leads to larger luminosities and bluer optical colors at early
times. Even with vigorous mixing it is difficult at very early times, $\ll
1$\,day after explosion, to explain very large luminosities or blue colors,
because the $^{56}$Ni has not had sufficient time to radioactively decay to
$^{56}$Co.} If the mixing is weak and the $^{56}$Ni is confined to the
innermost layers of the ejecta, the dark phase can last several days. Weak
mixing results in redder colors and a more moderate rise in luminosity
\citep{2014MNRAS.441..532D,2016ApJ...826...96P}. Thus, the early light
curves of even nonexotic SNe Ia convey information about their progenitor
systems and explosion mechanisms by constraining the distribution of
$^{56}$Ni.

@@

Using the SuperNova IDentification (\texttt{SNID};
\citealt{2007ApJ...666.1024B}) package, we find that the low-resolution
spectrum of \abc\ at $+18.8$\,days is best matched by normal SNe Ia. Several
characteristic features of an SN Ia, such as \ion{Si}{2} and \ion{S}{2}, can
be easily identified in \abc\ (Figure \ref{fig:spec_seq}). From the $+3.7 \,
\mathrm{day}$ LCO spectrum, we measure the pseudo-equivalent widths (pEWs)
of $-12 \pm 2$\,\AA\ and $-55 \pm 5$\,\AA\ for the absorption features near
5750 and 6100 \AA, attributed to \ion{Si}{2}\,$\lambda\lambda$5972, 6355,
respectively. According to the \citet{2006PASP..118..560B} classification
scheme, \abc\ is a shallow-silicon SN, similar to 1999aa-like SNe
\citep{2009PASP..121..238B}. In Figure~\ref{fig:branch_vel, the velocity
evolution of the \abc\ \ion{Si}{2}$\,\lambda6355$ absorption minimum is
compared to the median evolution of the four spectroscopic subclasses from
\citet{2006PASP..118..560B}. The median evolution is defined using the
sample of SNe Ia in \citet{2012AJ....143..126B}. For each SN, we interpolate
the \ion{Si}{2} velocity, $v_{\mathrm{Si\,II}\,\lambda6355}$, to a fixed
grid at 1\,day intervals. The curves are defined by the median
$v_{\mathrm{Si\,II}\,\lambda6355}$ at each point on the grid with at least
three SNe (this prevents just one or two SNe from defining the evolution of
an entire subclass). The velocity evolution of \abc\ is most reminiscent of
the shallow-silicon subclass.}

\begin{figure}[htb]
  \centering
  \includegraphics[width=3.35in]{Branch_vel_evol.pdf}
  \caption{
  Velocity evolution of \ion{Si{2},
  $v_{\mathrm{Si\,II}\,\lambda6355}$, for \abc\ compared to the median
  evolution of the four spectroscopic subclasses defined in
  \citet{2006PASP..118..560B}, using data from \citet{2012AJ....143..126B}.
  Typical uncertainties for \abc\ are $\sim$1000\,km\,s$^{-1}$ before
  $T_{B,\mathrm{max}}$, when the \ion{Si}{2}$\,\lambda6355$ profile is shallow
  and the minimum of absorption is difficult to determine, and
  $\sim$300\,km\,s$^{-1}$ after $T_{B,\mathrm{max}}$. For the median curves,
  the typical scatter, determined via the interquartile range of the sample,
  is $\sim$700\,km\,s$^{-1}$, around $T_{B,\mathrm{max}}$. At early times,
  core-normal and broad-line SNe have significantly faster \ion{Si}{2} than
  \abc, while the declining trend of cool SNe does not match \abc.} }
  \label{fig:branch_vel}
\end{figure}

@@

We measure the (pseudo-)bolometric luminosity, $L_\mathrm{UVOIR}$, of \abc\
at peak via trapezoidal integration of the reddening-corrected flux from the
UV, optical, and near-IR (UVOIR) filters. The light curves in the individual
filters are interpolated so that $L_\mathrm{UVOIR}$ is evaluated at common
epochs in each filter. From this integration, we measure a maximum
luminosity $L_\mathrm{max} = 1.2 \pm 0.1 \times 10^{43} \; \mathrm{erg \;
s}^{-1}$ for \abc. This value is consistent with the normal SNe Ia studied
in \citet{2016A&A...588A..84D}. Following Arnett's rule
\citep{1982ApJ...253..785A, 1985Natur.314..337A}, the mass of $^{56}$Ni
synthesized in the explosion can be derived from $L_\mathrm{max}$. Assuming
a rise time of $19 \pm 3$\,days (see \citealt{2006A&A...450..241S}), we find
$M_\mathrm{Ni} = 0.6 \pm 0.1 M_\odot$.\footnote{A 17.9\,day rise time
(\S\ref{sec:lc_fit}), yields a consistent estimate of $M_\mathrm{Ni}$.}}

@@

The time of explosion $t_\mathrm{exp}$ is not equal to $t_0$ (see above);
thus, \citet{2014ApJ...784...85P} suggest that measurements of the
photospheric velocity can determine $t_\mathrm{exp}$ given that the ejecta
begin expanding from the moment of explosion. Assuming a constant opacity in
the ejecta, \citeauthor{2014ApJ...784...85P} find that the photospheric
velocity evolves as $v_\mathrm{ph}\propto(t-t_\mathrm{exp})^{-0.22}$.
Numerical experiments by \citet{2016ApJ...826...96P} find that the
constant-opacity assumption strongly depends on the amount of $^{56}$Ni
mixing in the SN ejecta. As a result, the adoption of a $t^{-0.22}$
power-law model may not be valid for all SNe Ia. Nevertheless, we proceed on
the assumption that \abc\ experienced strong $^{56}$Ni mixing (see
\S\ref{sec:Ni_mixing)}, corresponding to the models that are best
approximated as a $t^{-0.22}$ power law. We do this in part to compare with
previous studies, though we caution that the inferred value of
$t_\mathrm{exp}$ is subject to uncertainties related to ejecta mixing.

@@

\begin{figure}[]
  \centering
  \includegraphics[width=3.35in]{compare_Noebauer.pdf}
  \caption{
  %
  Comparison of the models from \citet{2017MNRAS.472.2787N to the \abc\
  $V$-band light curve (\textit{top}) and the $(B-V)_0$ color curve
  (\textit{bottom}). The light curve has been corrected for the distance
  modulus to \abc\ (\S\ref{sec:host}), while both the light curve and color
  curve have been corrected for reddening. To guide the eye,
  $g_\mathrm{PTF}$ observations are shown, though we caution that \abc\ may
  exhibit significant color evolution at this phase, in which case
  $g_\mathrm{PTF}$ would be a poor proxy for $V$. Each model light curve is
  translated to match the LCO observations $\sim$1.7\,days after $t_0$ given
  the uncertain time of explosion. Translational offsets are listed in the
  bottom panel legend, which shows the models in order of decreasing
  $^{56}$Ni from top to bottom, then left to right. The inset in the top
  panel shows the residuals relative to the W7 model.} }
  %
  \label{fig:noebauer} 
\end{figure}

@@

Figure~\ref{fig:piro} compares the models from \citet{2016ApJ...826...96P}
to \abc. The \citeauthor{2016ApJ...826...96P} models employ a piston-driven
explosion to explode a single WD progenitor model. As the piston explosion
does not result in any nucleosynthesis, the distribution of $^{56}$Ni in the
ejecta must be prescribed by hand, which enables a study of the effects of
mixing on the resulting SN emission. Each model employs a fixed $0.5\,\sm$
of $^{56}$Ni that has been distributed throughout the ejecta via boxcar
averaging (see their Figure~1). The resulting light curves are synthesized
using the SuperNova Explosion Code (\texttt{SNEC};
\citealt{2015ApJ...814...63M}), as shown in Figure~\ref{fig:piro}. Broadly
speaking, the results can be summarized as follows: SNe with strong mixing
exhibit a rapid rise and quickly develop blue colors, whereas models where
the $^{56}$Ni is confined to the innermost layers of the ejecta remain faint
for days after explosion and feature a gradual color evolution from the red
to the blue. The model with the strongest mixing (dark long-dashed line in
Figure~\ref{fig:piro)} best matches the observations of \abc. This is the
only model we have found that exhibits a flat $(B-V)_0$ color evolution in
the days after explosion; however, in detail this model is too red relative
to the observations of \abc.

>>>>

%% Reintroduced the \received and \accepted commands from AASTeX v5.2
\received{2017 August 23}
\revised{2017 November 23}
\accepted{2017 December 5}
\published{2018 January 11}
%% Command to document which AAS Journal the manuscript was submitted to.
%% Adds "Submitted to " the arguement.
\submitjournal{ApJ}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% The following section outlines numerous optional output that
%% can be displayed in the front matter or as running meta-data.
%%
%% If you wish, you may supply running head information, although
%% this information may be modified by the editorial offices.
\shorttitle{\abc}
\shortauthors{Miller et al.}
%%
%% You can add a light gray and diagonal water-mark to the first page 
%% with this command:
% \watermark{DRAFT}
%% where "text", e.g. DRAFT, is the text to appear.  If the text is 
%% long you can control the water-mark size with:
%  \setwatermarkfontsize{dimension}
%% where dimension is any recognized LaTeX dimension, e.g. pt, in, etc.
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Early observations of Type Ia supernovae (SNe Ia) provide a unique probe of
their progenitor systems and explosion physics. Here we report the
intermediate Palomar Transient Factory (iPTF) discovery of an
extraordinarily young SN Ia, \abc. By fitting a power law to our early light
curve, we infer that first light for the SN, that is when the SN could have
first been detected by our survey, occurred only
$0.15\pm_{0.07}^{0.15}$\,days before our first detection. In the
$\sim$24\,hr after discovery, \abc\ rose by $\sim$2\,mag, featuring a
near-linear rise in flux for $\gtrsim$3\,days. Early spectra show strong
\ion{C}{2} absorption, which disappears after $\sim$7\,days. Unlike the
extensivelyobserved SN Ia SN\,2011fe, the $(B-V)_0$ colors of \abc\ are blue
and nearly constant in the days after explosion. We show that our early
observations of \abc\ cannot be explained by either SN shock breakout and
the associated, subsequent cooling or the SN ejecta colliding with a stellar
companion. Instead, we argue that the early characteristics of \abc,
including (i) the rapid, near-linear rise, (ii) the nonevolving blue colors,
and (iii) the strong \ion{C}{2} absorption, are the result of either ejecta
interaction with nearby, unbound material or vigorous mixing of radioactive
$^{56}$Ni in the SN ejecta, or a combination of the two. In the next few
years, dozens of very young \textit{normal} SNe Ia will be discovered, and
observations similar to those presented here will constrain the white dwarf
explosion mechanism.

@@

The vast majority of SNe Ia are observed to be powered purely by the
radioactive decay of $^{56}$Ni. While the detection of SN shock cooling or
ejecta interaction is rare, the level of $^{56}$Ni mixing in the SN ejecta
can fundamentally alter the appearance of the SN shortly after explosion
(e.g.,
\citealt{2014MNRAS.441..532D,2016ApJ...826...96P,2017MNRAS.472.2787N}).

SNe Ia experience a dark phase after the SN shock breakout but before
radioactive energy diffuses into the photosphere
\citep{2014ApJ...784...85P}. The duration of this dark phase is set by how
the newly synthesized $^{56}$Ni is mixed and deposited into different layers
of the ejecta. Strong mixing leads to a short, or nonexistent, dark phase,
because the radioactive $\gamma$-rays rapidly diffuse to the photosphere.
This also leads to larger luminosities and bluer optical colors at early
times. Even with vigorous mixing it is difficult at very early times, $\ll
1$\,day after explosion, to explain very large luminosities or blue colors,
because the $^{56}$Ni has not had sufficient time to radioactively decay to
$^{56}$Co. If the mixing is weak and the $^{56}$Ni is confined to the
innermost layers of the ejecta, the dark phase can last several days. Weak
mixing results in redder colors and a more moderate rise in luminosity
\citep{2014MNRAS.441..532D,2016ApJ...826...96P}. Thus, the early light
curves of even nonexotic SNe Ia convey information about their progenitor
systems and explosion mechanisms by constraining the distribution of
$^{56}$Ni.

@@

Using the SuperNova IDentification (\texttt{SNID};
\citealt{2007ApJ...666.1024B}) package, we find that the low-resolution
spectrum of \abc\ at $+18.8$\,days is best matched by normal SNe Ia. Several
characteristic features of an SN Ia, such as \ion{Si}{2} and \ion{S}{2}, can
be easily identified in \abc\ (Figure \ref{fig:spec_seq}). From the $+3.7 \,
\mathrm{day}$ LCO spectrum, we measure the pseudo-equivalent widths (pEWs)
of $-12 \pm 2$\,\AA\ and $-55 \pm 5$\,\AA\ for the absorption features near
5750 and 6100 \AA, attributed to \ion{Si}{2}\,$\lambda\lambda$5972, 6355,
respectively. According to the \citet{2006PASP..118..560B} classification
scheme, \abc\ is a shallow-silicon SN, similar to 1999aa-like SNe
\citep{2009PASP..121..238B}. In Figure~\ref{fig:branch_vel}, the velocity
evolution of the \abc\ \ion{Si}{2}$\,\lambda6355$ absorption minimum is
compared to the median evolution of the four spectroscopic subclasses from
\citet{2006PASP..118..560B}. The median evolution is defined using the
sample of SNe Ia in \citet{2012AJ....143..126B}. For each SN, we interpolate
the \ion{Si}{2} velocity, $v_{\mathrm{Si\,II}\,\lambda6355}$, to a fixed
grid at 1\,day intervals. The curves are defined by the median
$v_{\mathrm{Si\,II}\,\lambda6355}$ at each point on the grid with at least
three SNe (this prevents just one or two SNe from defining the evolution of
an entire subclass). The velocity evolution of \abc\ is most reminiscent of
the shallow-silicon subclass.

\begin{figure}[htb]
  \centering
  \includegraphics[width=3.35in]{Branch_vel_evol.pdf}
  \caption{
  Velocity evolution of \ion{Si}{2},
  $v_{\mathrm{Si\,II}\,\lambda6355}$, for \abc\ compared to the median
  evolution of the four spectroscopic subclasses defined in
  \citet{2006PASP..118..560B}, using data from \citet{2012AJ....143..126B}.
  Typical uncertainties for \abc\ are $\sim$1000\,km\,s$^{-1}$ before
  $T_{B,\mathrm{max}}$, when the \ion{Si}{2}$\,\lambda6355$ profile is shallow
  and the minimum of absorption is difficult to determine, and
  $\sim$300\,km\,s$^{-1}$ after $T_{B,\mathrm{max}}$. For the median curves,
  the typical scatter, determined via the interquartile range of the sample,
  is $\sim$700\,km\,s$^{-1}$, around $T_{B,\mathrm{max}}$. At early times,
  core-normal and broad-line SNe have significantly faster \ion{Si}{2} than
  \abc, while the declining trend of cool SNe does not match \abc.}
  \label{fig:branch_vel}
\end{figure}

@@

We measure the (pseudo-)bolometric luminosity, $L_\mathrm{UVOIR}$, of \abc\
at peak via trapezoidal integration of the reddening-corrected flux from the
UV, optical, and near-IR (UVOIR) filters. The light curves in the individual
filters are interpolated so that $L_\mathrm{UVOIR}$ is evaluated at common
epochs in each filter. From this integration, we measure a maximum
luminosity $L_\mathrm{max} = 1.2 \pm 0.1 \times 10^{43} \; \mathrm{erg \;
s}^{-1}$ for \abc. This value is consistent with the normal SNe Ia studied
in \citet{2016A&A...588A..84D}. Following Arnett's rule
\citep{1982ApJ...253..785A, 1985Natur.314..337A}, the mass of $^{56}$Ni
synthesized in the explosion can be derived from $L_\mathrm{max}$. Assuming
a rise time of $19 \pm 3$\,days (see \citealt{2006A&A...450..241S}), we find
$M_\mathrm{Ni} = 0.6 \pm 0.1 M_\odot$.\footnote{A 17.9\,day rise time
(\S\ref{sec:lc_fit}), yields a consistent estimate of $M_\mathrm{Ni}$.}

@@

The time of explosion $t_\mathrm{exp}$ is not equal to $t_0$ (see above);
thus, \citet{2014ApJ...784...85P} suggest that measurements of the
photospheric velocity can determine $t_\mathrm{exp}$ given that the ejecta
begin expanding from the moment of explosion. Assuming a constant opacity in
the ejecta, \citeauthor{2014ApJ...784...85P} find that the photospheric
velocity evolves as $v_\mathrm{ph}\propto(t-t_\mathrm{exp})^{-0.22}$.
Numerical experiments by \citet{2016ApJ...826...96P} find that the
constant-opacity assumption strongly depends on the amount of $^{56}$Ni
mixing in the SN ejecta. As a result, the adoption of a $t^{-0.22}$
power-law model may not be valid for all SNe Ia. Nevertheless, we proceed on
the assumption that \abc\ experienced strong $^{56}$Ni mixing (see
\S\ref{sec:Ni_mixing}), corresponding to the models that are best
approximated as a $t^{-0.22}$ power law. We do this in part to compare with
previous studies, though we caution that the inferred value of
$t_\mathrm{exp}$ is subject to uncertainties related to ejecta mixing.

@@

\begin{figure}[]
  \centering
  \includegraphics[width=3.35in]{compare_Noebauer.pdf}
  \caption{
  %
  Comparison of the models from \citet{2017MNRAS.472.2787N} to the \abc\
  $V$-band light curve (\textit{top}) and the $(B-V)_0$ color curve
  (\textit{bottom}). The light curve has been corrected for the distance
  modulus to \abc\ (\S\ref{sec:host}), while both the light curve and color
  curve have been corrected for reddening. To guide the eye,
  $g_\mathrm{PTF}$ observations are shown, though we caution that \abc\ may
  exhibit significant color evolution at this phase, in which case
  $g_\mathrm{PTF}$ would be a poor proxy for $V$. Each model light curve is
  translated to match the LCO observations $\sim$1.7\,days after $t_0$ given
  the uncertain time of explosion. Translational offsets are listed in the
  bottom panel legend, which shows the models in order of decreasing
  $^{56}$Ni from top to bottom, then left to right. The inset in the top
  panel shows the residuals relative to the W7 model.}
  %
  \label{fig:noebauer} 
\end{figure}

@@

Figure~\ref{fig:piro} compares the models from \citet{2016ApJ...826...96P}
to \abc. The \citeauthor{2016ApJ...826...96P} models employ a piston-driven
explosion to explode a single WD progenitor model. As the piston explosion
does not result in any nucleosynthesis, the distribution of $^{56}$Ni in the
ejecta must be prescribed by hand, which enables a study of the effects of
mixing on the resulting SN emission. Each model employs a fixed $0.5\,\sm$
of $^{56}$Ni that has been distributed throughout the ejecta via boxcar
averaging (see their Figure~1). The resulting light curves are synthesized
using the SuperNova Explosion Code (\texttt{SNEC};
\citealt{2015ApJ...814...63M}), as shown in Figure~\ref{fig:piro}. Broadly
speaking, the results can be summarized as follows: SNe with strong mixing
exhibit a rapid rise and quickly develop blue colors, whereas models where
the $^{56}$Ni is confined to the innermost layers of the ejecta remain faint
for days after explosion and feature a gradual color evolution from the red
to the blue. The model with the strongest mixing (dark long-dashed line in
Figure~\ref{fig:piro}) best matches the observations of \abc. This is the
only model we have found that exhibits a flat $(B-V)_0$ color evolution in
the days after explosion; however, in detail this model is too red relative
to the observations of \abc.
